MCR_Algae_Data_Cleaning

### 2024-09-24 Personal_Prelim_Code/julianna_figs_Aug29.Rmd
I HAVE MOVED YOUR PERSONAL CODE.
I did not delete it, but it is in a new folder. your file paths may not be happy, but the code is still there.

```{r get a dataset with the main four spaceholders}
macroalgQuad %>% 
  select(year, location, site, habitat, transect, quadrat, island_side, macroalgae_cover, macroalgae_cover_hard, coral_cover, coral_cover_hard, sand_cover) %>% 
  full_join(algae) %>%
  rowwise() %>% 
  # get turf and CCA cover
  mutate(turf_cover = sum(algal_turf, damselfish_turf),
         turf_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (turf_cover/(100-sand_cover))*100 
           ),
         CCA_cover = sum(lithophyllum_kotschyanum, crustose_corallines, neogoniolithon_brassica_florida, sporolithon_sp),
         CCA_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (CCA_cover/(100-sand_cover))*100 
           )
         ) %>% 
  select(year, location, site, habitat, transect, quadrat, island_side, 
         macroalgae_cover, macroalgae_cover_hard, coral_cover, coral_cover_hard, sand_cover, 
         turf_cover, turf_cover_hard, CCA_cover, CCA_cover_hard
         ) %>% 
  # relevel for graphing
  mutate(island_side = str_to_sentence(island_side),
         island_side = factor(island_side, levels = c("North", "East", "West")),
         habitat = case_when(
           str_detect(habitat, pattern = "_") ~ str_to_sentence(paste0(str_split(habitat, "_")[[1]][1], " ", str_split(habitat, "_")[[1]][2])),
           TRUE ~ str_to_sentence(habitat)),
         habitat = factor(habitat, levels = c("Fringing", "Backreef", "Outer 10", "Outer 17"))) %>% 
  mutate(site = case_when(
    site == "lter_1" ~ "LTER 1 (N)",
    site == "lter_2" ~ "LTER 2 (N)",
    site == "lter_3" ~ "LTER 3 (E)",
    site == "lter_4" ~ "LTER 4 (E)",
    site == "lter_5" ~ "LTER 5 (W)",
    site == "lter_6" ~ "LTER 6 (W)"
  )) -> main4quad
```

```{r plot for absolute cover}
main4quad %>% 
  select(year, location, site, habitat, transect, quadrat, island_side, 
         macroalgae_cover, turf_cover, CCA_cover, coral_cover) %>% 
  # pivot key columns longer
  pivot_longer(cols = c(macroalgae_cover, turf_cover, CCA_cover, coral_cover),
               names_to = "Substrate",
               values_to = "cover") %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, site, island_side, Substrate) %>% 
  summarise(mean_cov = mean(cover),
            sd_cov = sd(cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # plot
  ggplot(aes(x = year, y = mean_cov, group = Substrate, fill = Substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
   scale_fill_manual(values = c(
    moma.colors("Klein", type = "discrete")[2], # CCA
    moma.colors("Klein", type = "discrete")[1], # coral
    moma.colors("Klein", type = "discrete")[5], # macro
    moma.colors("Klein", type = "discrete")[4] # turf
  ),
  labels = c("CCA", "Coral", "Macroalgae", "Turf algae")) +
  facet_grid(site ~ habitat) + # not sure which direction is best here
  ylab("Mean absolute cover (%)") +
  xlab("Year") +
  # add lines for disturbances
  geom_vline(xintercept = 2006, linetype = "dotdash") + # start of COTS
  geom_vline(xintercept = 2010,  color = "darkgray") + # cyclone Oli
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red") 
```

>>>>

```{r get a dataset with the main four spaceholders}
macroalgQuad %>% 
  select(year, location, site, habitat, transect, quadrat, island_side, macroalgae_cover, macroalgae_cover_hard, coral_cover, coral_cover_hard, sand_cover) %>% 
  full_join(algae) %>%
  rowwise() %>% 
  # get turf and CCA cover
  mutate(turf_cover = sum(algal_turf, damselfish_turf),
         turf_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (turf_cover/(100-sand_cover))*100 
           ),
         CCA_cover = sum(lithophyllum_kotschyanum, crustose_corallines, neogoniolithon_brassica_florida, sporolithon_sp),
         CCA_cover_hard = case_when(
           sand_cover == 100 ~ NA, # since we can't divide by a zero
           TRUE ~ (CCA_cover/(100-sand_cover))*100 
           )
         ) %>% 
  select(year, location, site, habitat, transect, quadrat, island_side, 
         macroalgae_cover, macroalgae_cover_hard, coral_cover, coral_cover_hard, sand_cover, 
         turf_cover, turf_cover_hard, CCA_cover, CCA_cover_hard, sand_cover
         ) %>% 
  # relevel for graphing
  mutate(island_side = str_to_sentence(island_side),
         island_side = factor(island_side, levels = c("North", "East", "West")),
         habitat = case_when(
           str_detect(habitat, pattern = "_") ~ str_to_sentence(paste0(str_split(habitat, "_")[[1]][1], " ", str_split(habitat, "_")[[1]][2])),
           TRUE ~ str_to_sentence(habitat)),
         habitat = factor(habitat, levels = c("Fringing", "Backreef", "Outer 10", "Outer 17"))) %>% 
  mutate(site = case_when(
    site == "lter_1" ~ "LTER 1 (N)",
    site == "lter_2" ~ "LTER 2 (N)",
    site == "lter_3" ~ "LTER 3 (E)",
    site == "lter_4" ~ "LTER 4 (E)",
    site == "lter_5" ~ "LTER 5 (W)",
    site == "lter_6" ~ "LTER 6 (W)"
  )) -> main4quad
```

```{r plot for absolute cover}
main4quad %>% 
  select(year, location, site, habitat, transect, quadrat, island_side, 
         macroalgae_cover, turf_cover, CCA_cover, coral_cover, sand_cover) %>% 
  # pivot key columns longer
  pivot_longer(cols = c(macroalgae_cover, turf_cover, CCA_cover, coral_cover, sand_cover),
               names_to = "Substrate",
               values_to = "cover") %>% 
  # summarize and group quadrats by island/habitat
  group_by(year, habitat, site, island_side, Substrate) %>% 
  summarise(mean_cov = mean(cover),
            sd_cov = sd(cover, na.rm = TRUE),
            n = n(),
            CI95 = 1.96*(sd_cov)/sqrt(n)) %>% 
  # plot
  ggplot(aes(x = year, y = mean_cov, group = Substrate, fill = Substrate)) +
  geom_ribbon(aes(ymin = mean_cov -CI95, 
                    ymax = mean_cov + CI95),
              alpha  = 0.7) +
  geom_line() +
  theme_bw() +
  # scale_fill_manual(values = c(
  #  moma.colors("Klein", type = "discrete")[2], # CCA
  #  moma.colors("Klein", type = "discrete")[1], # coral
  #  moma.colors("Klein", type = "discrete")[5], # macro
  #  moma.colors("Klein", type = "discrete")[4] # turf
  #),
  #labels = c("CCA", "Coral", "Macroalgae", "Turf algae")) +
  facet_grid(site ~ habitat) + # not sure which direction is best here
  ylab("Mean absolute cover (%)") +
  xlab("Year") +
  # add lines for disturbances
  geom_vline(xintercept = 2006, linetype = "dotdash") + # start of COTS
  geom_vline(xintercept = 2010,  color = "darkgray") + # cyclone Oli
  geom_vline(xintercept = 2019, linetype = "dashed", color = "red") 
```

### 2024-09-24 Personal_Prelim_Code/mcr_algal_data_cleaning_final.Rmd
I HAVE MOVED YOUR PERSONAL CODE.
I did not delete it, but it is in a new folder. your file paths may not be happy, but the code is still there.

wide_alldata_filtered_cleaned_column_names <- 
  wide_alldata_filtered %>%
  janitor::clean_names()

# Julianna is adding this code here IF we want it:
wide_alldata_filtered_cleaned_column_names %>% 
  mutate(location = gsub(" ", "_", str_to_lower(location)),
         site = gsub(" ", "_", str_to_lower(site)),
         habitat = gsub(" ", "_", str_to_lower(habitat)),
         site_habitat = gsub(" ", "_", str_to_lower(site_habitat)),
         ) # would want to save this then do it for the algae dataset as well

write.csv(wide_alldata_filtered_cleaned_column_names, "allbenthicdata_cleaned_colnames.csv", row.names = FALSE)

@@

wide_algaeonly_filtered <- wide_alldata_filtered[ , -which(names(wide_alldata_filtered) 
                                                           %in% not_algae)]

@@

wide_algaeonly_cleaned_column_names <- 
  wide_algaeonly_filtered %>%
  janitor::clean_names()

write.csv(wide_algaeonly_cleaned_column_names, "algaeonly_cleaned_colnames.csv", row.names = FALSE)

>>>>

wide_alldata_filtered_cleaned_column_names <- 
  wide_alldata_filtered %>%
  janitor::clean_names() %>% 
  # Julianna is adding this code here IF we want it to clean the inside:
  mutate(location = gsub(" ", "_", str_to_lower(location)),
         site = gsub(" ", "_", str_to_lower(site)),
         habitat = gsub(" ", "_", str_to_lower(habitat)),
         site_habitat = gsub(" ", "_", str_to_lower(site_habitat)),
         )

write.csv(wide_alldata_filtered_cleaned_column_names, "data/allbenthicdata_cleaned_colnames.csv", row.names = FALSE)

@@

wide_algaeonly_filtered <- wide_alldatsa_filtered[ , -which(names(wide_alldata_filtered) 
                                                           %in% not_algae)]

@@

wide_algaeonly_cleaned_column_names <- 
  wide_algaeonly_filtered %>%
  # clean column names
  janitor::clean_names() %>% 
  # clean the inside as well
  mutate(location = gsub(" ", "_", str_to_lower(location)),
         site = gsub(" ", "_", str_to_lower(site)),
         habitat = gsub(" ", "_", str_to_lower(habitat)),
         site_habitat = gsub(" ", "_", str_to_lower(site_habitat)),
         )

write.csv(wide_algaeonly_cleaned_column_names, "data/algaeonly_cleaned_colnames.csv", row.names = FALSE)
